---
title: "analysis"
author: "Olivia Webster"
date: "2026-01-24"
output: html_document
---


#INTRODUCTION
thank you for being here!! this is my first project and I'm excited to show you!

##The Data
- This data is taken from the CDC National Health and Nutrition Examination Survery (https://wwwn.cdc.gov/nchs/nhanes/Default.aspx).
- I took two Continuous Studies from NHANES, the "NHANES 08/2021-08/2023" study and the "NHANES 2017-March 2020 Pre-Pandemic Data" study
    - **hereafter, the "NHANES 08/2021-08/2023" study will be referred as the "post-pandemic study", and the "NHANES 2017-March 2020 Pre-Pandemic Data" study will be referred to as the "pre-pandemic study".**
- within the pre-pandemic study, I extracted one (1) dataset from the Examination Data, "Blood Pressure - Oscillometric Measurements", and two (2) datasets within the post-pandemic study: "Blood Pressure - Oscillometric Measurements" from the Examination Data, and "Demographic Variables and Sample Weights" from the Demographics Data.

##The Process
- after graduating from William & Mary in December 2025, I wanted to apply my statistical knowledge to something I was interested in, like health-related subjects.
- In the beginning of planning to develop this project, I planned to answer 2 questions:
1. whether blood pressure (diastolic/systolic) decreased/increased/stayed the same before vs after the COVID-19 pandemic; aka blood pressure decreased/increased/stayed the same in the pre-pandemic study vs the post-pandemic study
2. if there is a relationship between the age/race/income/level of education of ADULT participants and their blood pressure reading.

### The Realization/Pivot
- I finished the t-tests and the age vs. diastolic bp in mid May, and was super happy with the results
- But after reading about how NHANES is not a completely randomized study, but stratified and clustered, I realized that I was not able to accurately answer the two questions without learning how to statistically analyze studies like this.

- with this in mind, I have decided to keep this version as an **exploratory, unweighted** analysis to practice only what I have learned in my statistical data analysis college courses, and once my knowledge about how to effectively analyze data sampled like this has expanded, I will develop a new version.


#THE CODE (woo!)

##Libraries and stuff
```{r setup, include=FALSE}

library(tidyverse)
install.packages("haven")
library(haven)
library(ggplot2)

##1
#blood pressure taken BEFORE COVID
file_path <- "C:/Users/oyaya/Downloads/NHANES_data/bp_2017.xpt"
before_bp <- read_xpt(file_path)
head(before_bp)
View(before_bp)


#bp of people in study AFTER COVID
file_path1 <- "C:/Users/oyaya/Downloads/NHANES_data/bp_2021.xpt"
after_bp <- read_xpt(file_path1)
summary(after_bp)
View(after_bp)
```



##SETUP for t-tests (filter data frames)

```{r}
##3 (setup)

#only show these columns


bbp <- before_bp |>
  select(SEQN, BPXOSY1:BPXODI3)

abp <- after_bp |>
  select(SEQN, BPXOSY1:BPXODI3)


after_bp <- after_bp |>
  mutate(BPXODI = (BPXODI1+BPXODI2+BPXODI3)/3, .before = BPXOSY1) |>
  mutate(BPXOSY = (BPXOSY1+BPXOSY2+BPXOSY3)/3, .before = BPXOSY1)

before_bp <- before_bp |>
  mutate(BPXODI = (BPXODI1+BPXODI2+BPXODI3)/3, .before = BPXOSY1) |>
  mutate(BPXOSY = (BPXOSY1+BPXOSY2+BPXOSY3)/3, .before = BPXOSY1)

```




##analyze blood pressure readings from BOTH studies
###graph each one

```{r}


hist(before_bp$BPXOSY, nclass=500, main = "systolic blood pressure before", xlab="systolic bp")

hist(after_bp$BPXOSY, nclass=500, main = "systolic blood pressure after", xlab="systolic bp")


hist(before_bp$BPXODI, nclass=500, main = "diastolic blood pressure before", xlab="diastolic bp")

hist(after_bp$BPXODI, nclass=500, main = "diastolic blood pressure after", xlab="diastolic bp")


```



###now descriptive statistics (summary, and the percentage of participants with high levels)

```{r}


summary(before_bp$BPXOSY)
mean(before_bp$BPXOSY >= 130, na.rm = TRUE) * 100


summary(after_bp$BPXOSY)
mean(after_bp$BPXOSY >= 130, na.rm = TRUE) * 100


summary(before_bp$BPXODI)
mean(before_bp$BPXODI >= 80, na.rm = TRUE) * 100

summary(after_bp$BPXODI)
mean(after_bp$BPXODI >= 80, na.rm = TRUE) * 100


```


##COMPARISONS
###t-tests comparing before vs after systolic blood pressure

Null hypothesis: mean(before) = mean(after)
alternative hypothesis: mean(before) !=/>/< mean(after)
```{r}

t.test(before_bp$BPXOSY, after_bp$BPXOSY, var.equal = FALSE)
t.test(before_bp$BPXOSY, after_bp$BPXOSY, alternative = "greater", var.equal = FALSE)

t.test(before_bp$BPXOSY)

t.test(after_bp$BPXOSY)

```

ok, so before is **greater than** after significantly woohoo! p = 0.006172


95% confidence for before_bp: [119.4326, 120.1818]
this means that if we were to take many more samples with sample size n, then 95% of the calculated CIs will cover or capture the true mean of this population.

95% confidence for after_bp: [118.6871, 119.5082]
this means that if we were to take many more samples with sample size n, then 95% of the calculated CIs will cover or capture the true mean of this population.


###t-test comparing before vs. after diastolic blood pressure

Null hypothesis: mean(before) = mean(after)
alternative hypothesis: mean(before) !=/>/< mean(after)
```{r}

t.test(before_bp$BPXODI, after_bp$BPXODI, var.equal = FALSE)
t.test(before_bp$BPXODI, after_bp$BPXODI, alternative = "less", var.equal = FALSE)

t.test(before_bp$BPXODI)


t.test(after_bp$BPXODI)


```
ok, so the before is **less than** after significantly woohoo! p = 0.0002751


95% confidence of before_bp: [71.37614, 71.84070]
this means that if we were to take many more samples with sample size n, then 95% of the calculated CIs will cover or capture the true mean  of this population.

95% confidence: [71.96296, 72.48243]
this means that if we were to take many more samples with sample size n, then 95% of the calculated CIs will cover or capture the true mean  of this population.


## REGRESSION
### Setup the demographics for the post-pandemic participants

```{r, include=FALSE}
dfile_path1 <- "C:/Users/oyaya/Downloads/NHANES_data/demo_2021.xpt"
after_demo <- read_xpt(dfile_path1)
View(after_demo)

ad <- after_demo |>
  select(SEQN,RIDAGEYR,RIDRETH3,DMDEDUC2)
```



### Combine demographics and blood pressure from the post-pandemic study to make it easier to analyze from one data frame 

```{r}
nhanes1 <- inner_join(ad, abp, by = 'SEQN')
View(nhanes1)

nhanes1 <- nhanes1 |>
  mutate(BPXODI = (BPXODI1+BPXODI2+BPXODI3)/3, .before = BPXOSY1) |>
  mutate(BPXOSY = (BPXOSY1+BPXOSY2+BPXOSY3)/3, .before = BPXOSY1)

```




### Plot Race vs Diastolic BP, Race vs Systolic BP, Age vs Diastolic BP, Age vs Systolic BP
```{r}
plot(nhanes1$RIDRETH3, nhanes1$BPXODI)
#Mexican-American has the lowest dbp

plot(nhanes1$RIDRETH3, nhanes1$BPXOSY)



#age vs dbp
plot(nhanes1$RIDAGEYR, nhanes1$BPXODI)
#looks quadratic; peaks at 40-60yrs


#age vs sbp
plot(nhanes1$RIDAGEYR, nhanes1$BPXOSY)
#linear??
```



### try MANY models
```{r}
bp.model <- lm(BPXODI ~ RIDAGEYR, data = nhanes1)
summary(bp.model)
#multiple R-sq = 0.09029

bp.model1 <- lm(I(BPXODI^2) ~ RIDAGEYR, data = nhanes1)
summary(bp.model1)
#multiple R-sq = 0.08201


bp.model2 <- lm(BPXODI ~ RIDAGEYR + I(RIDAGEYR^2), data = nhanes1)
summary(bp.model2)
#multiple R-sq = 0.2291

bp.model3 <- lm(BPXODI ~ RIDAGEYR + I(RIDAGEYR^2) + I(RIDAGEYR^3), data = nhanes1)
summary(bp.model3)
#multiple R-sq = 0.2291



bp.model5 <- lm(BPXODI ~ RIDAGEYR + I(RIDAGEYR^2) + I(RIDAGEYR^3) + I(RIDAGEYR^4) + I(RIDAGEYR^5), data = nhanes1)
summary(bp.model5)
#multiple R-sq = 0.2298

bp.model9 <- lm(BPXODI ~ RIDAGEYR + I(RIDAGEYR^2) + I(RIDAGEYR^3) + I(RIDAGEYR^4) + I(RIDAGEYR^5) + I(RIDAGEYR^6) + I(RIDAGEYR^7) + I(RIDAGEYR^8) + I(RIDAGEYR^9), data = nhanes1)
summary(bp.model9)
#multiple R-sq = 0.2301

x2 = as.numeric(nhanes1$RIDAGEYR>48)
x2_star = (nhanes1$RIDAGEYR-48)*x2
pw.bp.model <- lm(BPXODI ~ RIDAGEYR + x2_star, data = nhanes1)
summary(pw.bp.model)
#multiple R-sq = 0.2258

anova(bp.model5, bp.model9)

nl_model1 = lm((BPXODI) ~ RIDAGEYR + I(RIDAGEYR^2), data = nhanes1)
summary(nl_model1) #R-sq = 0.09588

nl_model2 = lm(log(BPXODI) ~ RIDAGEYR + I(RIDAGEYR^2), data = nhanes1)
summary(nl_model2)
#R-sq (multiple) 0.2443
#we'll use this one!

```
We'll use the nl_model2


###Remove NAs
```{r}
nhanes1 <- nhanes1 |>
  filter(!is.na(BPXODI))
```


## PLOT MODEL AND DATA
```{r}
ggplot(nhanes1, aes(x = RIDAGEYR, y = log(BPXODI))) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              color = "blue",
              se = TRUE) +
  labs(title = "Quadratic Regression Fit",
       x = "Age",
       y = "log(BPXODI)")
```


## RESIDUAL ANALYSIS

```{r}
res <- nl_model2$residuals #got it!!


plot(fitted(nl_model2), res)

qqnorm(res)
qqline(res)
```

